Trans-Dimensional Bayesian Inference for Gravitational Lens Substructures

### 2015-09-07 Papers/ms.tex
Mention residuals
Filenames etc changed to make way for dataset 2

In the present paper we do not address the question of exactly which quantities
related to the
density profile of the lens are most scientifically interesting. Nevertheless,
we can verify that the results of the inference do behave in understandable
ways. For example, in Figure~\ref{fig:masses},
we plot the joint posterior distribution
for the SIE mass within its critical ellipse and the total substructure mass
over the entire domain. As one would expect, there is a strong dependence
between these two quantities in the posterior distribution, as mass in the
SIE can be traded off with mass in substructures to some extent, while
the model remains (loosely speaking) ``consistent with the data''.
\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{masses.pdf}
\caption{The joint posterior distribution for the SIE mass (integrated within
its critical ellipse, which is not the critical curve of the lens overall),
and the total mass in substructures.\label{fig:masses}}
\end{center}
\end{figure}

The posterior distribution for $N_{\rm lens}$ is also clearly of interest, and
is displayed in Figure~\ref{fig:N_lens}. The prior for this parameter was
uniform from 0 to 10 (inclusive), and the possibility $N=0$ has been
(loosely speaking) ``ruled out'' by the image data. The true solution ($N=1$)
has the highest probability. However, the possibilities with $N > 1$ are
still fairly plausible, since it's possible that a very small substructure
exists, and it's also possible that two substructures near each other could
mimic the effect of one. The degree to which these possibilities are plausible
is related to the choice of prior for the blob amplitudes and positions.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{N_lens.pdf}
\caption{The marginal posterior distribution for $N_{\rm lens}$, the
number of substructures in the lens. The prior was uniform and the true
value used to generate the data was 1. However, the data is only informative
enough to suppress the probability of values above 1 slightly, since it
is possible (given this data) that low mass substructures might exist somewhere,
or that what we think is one substructure might actually be two close together,
and other such possibilities.
\label{fig:N_lens}}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{substructures.pdf}
\caption{The {\revisions `easy'} simulated data, with substructure positions overlaid (a
Monte Carlo approximation to the posterior expectation of the empirical
measure). The density of black points in any region is proportional to the
expected number of substructures whose centers lie within that region. In this
case, there is strong evidence for a substructure close to the top-left image
(where one was actually placed).
\label{fig:substructures}}
\end{center}
\end{figure}

@@

The joint posterior distribution for the SIE mass total substructure mass
for the Cosmic Horseshoe is given in Figure~\ref{fig:masses2}. As with
the simulated data (Figure~\ref{fig:masses}), we see an expected negative
correlation between these two quantities. However, the points are not as
smoothly distributed in this case (we discuss this issue further
in Section~\ref{sec:convergence}). An upper ``limit'' on the SIE mass, with
95\% posterior probability, is $5.21 \times 10^{12}$ solar masses.
The algorithm has also found
some possibilities where the substructure mass is much greater than this. In
these cases, the substructures are far from the image --- it is the environment
that is being modelled.

The posterior distribution for
$N_{\rm lens}$ (Figure~\ref{fig:N_lens2})
also shows some evidence for more than zero substructures.
In particular, the prior probability for $N_{\rm lens} > 0$ was 10/11
$\approx$ 0.91, whereas the posterior probability is approximately 0.97.
This corresponds to a Bayes Factor of approximately 3 in favour of
$N_{\rm lens} > 0$ versus the alternative $N_{\rm lens} = 0$.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{masses2.pdf}
\caption{The joint posterior distribution for the SIE mass (integrated within
its critical ellipse, which is not the critical curve of the lens overall),
and the total mass in substructures. The units are defined by the critical
density, so that a mass of $\pi$ units would have an Einstein radius of one arcsecond. To calculate the masses in solar masses, we assumed a flat cosmology with
$\Omega_m=0.3$, $\Omega_\Lambda=0.7$, and
$H_0=70$ km/s/Mpc.
\label{fig:masses2}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{N_lens2.pdf}
\caption{The posterior distribution for the number of lens substructures
in the Cosmic Horseshoe system. There is mild evidence in favour of the
hypothesis that $N_{\rm lens} \neq 0$.\label{fig:N_lens2}}
\end{center}
\end{figure}

Despite weak evidence for the existence of substructure, when we examine the
expected value of the empirical measure of substructure positions
(Figure~\ref{fig:substructures2}) we find no consistency in their positions,
unlike for the simulated data (Figure~\ref{fig:substructures}). This situation
is not uncommon. For example, \citet{exoplanet} found evidence for a large
number of Keplerian signals in a time series, but the number of such signals
with well constrained properties was much lower. Related to this, we can
compute posterior probabilities for any hypothesis about the substructure
masses. With 95\% probability the substructure mass is less than 5.80$\times 10^{12}$ solar masses and with 75\% probability it is less than 3.56$\times 10^{12}$ solar masses. As the Einstein ring itself implies a mass of about $5 \times 10^{12}$
solar masses, these summaries are affected by the possibility of blobs outside
the main Einstein ring.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{substructures2.pdf}
\caption{The positions of substructures encountered in the MCMC sampling
for the Cosmic Horseshoe; technically a Monte Carlo approximation to the
posterior expected value of the empirical measure of substructure
positions. Unlike Figure~\ref{fig:substructures}, for the Cosmic Horseshoe
there is little consistency in the positions of the substructures.
\label{fig:substructures2}}
\end{center}
\end{figure}

The estimated marginal likelihood of our model
is
$\ln\left[p(\boldsymbol{D} | \boldsymbol{\theta})\right] \approx -6967$, and
the information (Kullback-Leibler divergence from the prior to the posterior)
is $\mathcal{H} \approx 129$ nats. 
Three example models (lens and source) sampled from the posterior are shown
in Figure~\ref{fig:sources}.

@@

\subsection{Reproducibility of the results}\label{sec:convergence}
The ``effective sample size'' returned by DNS, which we have described as the
number of posterior samples, takes into account the fact DNS's target
distribution is not the posterior. However, it does not take autocorrelation
into account, and thus can present an optimistic picture of the accuracy of
any Monte Carlo approximations to posterior quantities. Most standard diagnostic
techniques used in MCMC can be applied here. The simplest of these is a
check of reproducibility. If different runs yield substantially different
results, the MCMC output should be treated with caution. For example, in
Figure~\ref{fig:masses2}, there is a correlation between the mass attributed
to the SIE component and that attributed to substructures. However, the
distribution also appears somewhat ``lumpy'' or multimodal. This may
not be a feature of the actual posterior distribution, but could arise due to
imperfect sampling. Whereas a standard Metropolis sampler would move around
very slowly in the hypothesis space, DNS naturally spends a non-negligible fraction
of the time sampling the prior. Therefore, a particle exploring the parameter
space can ``forget'' its good fitting position, move somewhere completely
different, and find another good fitting model in a different location. This
is a natural feature of the algorithm (and is also present in related
algorithms such as parallel tempering). Despite this, the patchy
nature of Figure~\ref{fig:masses2} suggests that posterior exploration is
still challenging in this problem.

The results for the simulated data (Figure~\ref{fig:masses})
were less complex because of the definite existence of one substructure.
Its mass and position provide some information about the hyperparameters
$\boldsymbol{\alpha}_{\rm lens}$, restricting the probability that
high mass substructures exist far from the image.

>>>>

In the present paper we do not address the question of exactly which quantities
related to the
density profile of the lens are most scientifically interesting. Nevertheless,
we can verify that the results of the inference do behave in understandable
ways. For example, in Figure~\ref{fig:masses1},
we plot the joint posterior distribution
for the SIE mass within its critical ellipse and the total substructure mass
over the entire domain. As one would expect, there is a strong dependence
between these two quantities in the posterior distribution, as mass in the
SIE can be traded off with mass in substructures to some extent, while
the model remains (loosely speaking) ``consistent with the data''.
\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{masses1.pdf}
\caption{The joint posterior distribution for the SIE mass (integrated within
its critical ellipse, which is not the critical curve of the lens overall),
and the total mass in substructures.\label{fig:masses1}}
\end{center}
\end{figure}

The posterior distribution for $N_{\rm lens}$ is also clearly of interest, and
is displayed in Figure~\ref{fig:N_lens1}. The prior for this parameter was
uniform from 0 to 10 (inclusive), and the possibility $N=0$ has been
(loosely speaking) ``ruled out'' by the image data. The true solution ($N=1$)
has the highest probability. However, the possibilities with $N > 1$ are
still fairly plausible, since it's possible that a very small substructure
exists, and it's also possible that two substructures near each other could
mimic the effect of one. The degree to which these possibilities are plausible
is related to the choice of prior for the blob amplitudes and positions.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{N_lens1.pdf}
\caption{The marginal posterior distribution for $N_{\rm lens}$, the
number of substructures in the lens. The prior was uniform and the true
value used to generate the data was 1. However, the data is only informative
enough to suppress the probability of values above 1 slightly, since it
is possible (given this data) that low mass substructures might exist somewhere,
or that what we think is one substructure might actually be two close together,
and other such possibilities.
\label{fig:N_lens1}}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{substructures1.pdf}
\caption{The {\revisions `easy'} simulated data, with substructure positions overlaid (a
Monte Carlo approximation to the posterior expectation of the empirical
measure). The density of black points in any region is proportional to the
expected number of substructures whose centers lie within that region. In this
case, there is strong evidence for a substructure close to the top-left image
(where one was actually placed).
\label{fig:substructures1}}
\end{center}
\end{figure}

@@

The joint posterior distribution for the SIE mass total substructure mass
for the Cosmic Horseshoe is given in Figure~\ref{fig:masses3}. As with
the simulated data (Figure~\ref{fig:masses1}), we see an expected negative
correlation between these two quantities. However, the points are not as
smoothly distributed in this case (we discuss this issue further
in Section~\ref{sec:convergence}). An upper ``limit'' on the SIE mass, with
95\% posterior probability, is $5.21 \times 10^{12}$ solar masses.
The algorithm has also found
some possibilities where the substructure mass is much greater than this. In
these cases, the substructures are far from the image --- it is the environment
that is being modelled.

The posterior distribution for
$N_{\rm lens}$ (Figure~\ref{fig:N_lens3})
also shows some evidence for more than zero substructures.
In particular, the prior probability for $N_{\rm lens} > 0$ was 10/11
$\approx$ 0.91, whereas the posterior probability is approximately 0.97.
This corresponds to a Bayes Factor of approximately 3 in favour of
$N_{\rm lens} > 0$ versus the alternative $N_{\rm lens} = 0$.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{masses3.pdf}
\caption{The joint posterior distribution for the SIE mass (integrated within
its critical ellipse, which is not the critical curve of the lens overall),
and the total mass in substructures. The units are defined by the critical
density, so that a mass of $\pi$ units would have an Einstein radius of one arcsecond. To calculate the masses in solar masses, we assumed a flat cosmology with
$\Omega_m=0.3$, $\Omega_\Lambda=0.7$, and
$H_0=70$ km/s/Mpc.
\label{fig:masses3}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{N_lens3.pdf}
\caption{The posterior distribution for the number of lens substructures
in the Cosmic Horseshoe system. There is mild evidence in favour of the
hypothesis that $N_{\rm lens} \neq 0$.\label{fig:N_lens3}}
\end{center}
\end{figure}

Despite weak evidence for the existence of substructure, when we examine the
expected value of the empirical measure of substructure positions
(Figure~\ref{fig:substructures3}) we find no consistency in their positions,
unlike for the simulated data (Figure~\ref{fig:substructures1}). This situation
is not uncommon. For example, \citet{exoplanet} found evidence for a large
number of Keplerian signals in a time series, but the number of such signals
with well constrained properties was much lower. Related to this, we can
compute posterior probabilities for any hypothesis about the substructure
masses. With 95\% probability the substructure mass is less than 5.80$\times 10^{12}$ solar masses and with 75\% probability it is less than 3.56$\times 10^{12}$ solar masses. As the Einstein ring itself implies a mass of about $5 \times 10^{12}$
solar masses, these summaries are affected by the possibility of blobs outside
the main Einstein ring.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{substructures3.pdf}
\caption{The positions of substructures encountered in the MCMC sampling
for the Cosmic Horseshoe; technically a Monte Carlo approximation to the
posterior expected value of the empirical measure of substructure
positions. Unlike Figure~\ref{fig:substructures1}, for the Cosmic Horseshoe
there is little consistency in the positions of the substructures.
\label{fig:substructures3}}
\end{center}
\end{figure}

The estimated marginal likelihood of our model
is
$\ln\left[p(\boldsymbol{D} | \boldsymbol{\theta})\right] \approx -6967$, and
the information (Kullback-Leibler divergence from the prior to the posterior)
is $\mathcal{H} \approx 129$ nats. 
Three example models (lens and source) sampled from the posterior are shown
in Figure~\ref{fig:sources}. {\revisions
The model was able to fit the data down to the noise level.
}

@@

\subsection{Reproducibility of the results}\label{sec:convergence}
The ``effective sample size'' returned by DNS, which we have described as the
number of posterior samples, takes into account the fact DNS's target
distribution is not the posterior. However, it does not take autocorrelation
into account, and thus can present an optimistic picture of the accuracy of
any Monte Carlo approximations to posterior quantities. Most standard diagnostic
techniques used in MCMC can be applied here. The simplest of these is a
check of reproducibility. If different runs yield substantially different
results, the MCMC output should be treated with caution. For example, in
Figure~\ref{fig:masses3}, there is a correlation between the mass attributed
to the SIE component and that attributed to substructures. However, the
distribution also appears somewhat ``lumpy'' or multimodal. This may
not be a feature of the actual posterior distribution, but could arise due to
imperfect sampling. Whereas a standard Metropolis sampler would move around
very slowly in the hypothesis space, DNS naturally spends a non-negligible fraction
of the time sampling the prior. Therefore, a particle exploring the parameter
space can ``forget'' its good fitting position, move somewhere completely
different, and find another good fitting model in a different location. This
is a natural feature of the algorithm (and is also present in related
algorithms such as parallel tempering). Despite this, the patchy
nature of Figure~\ref{fig:masses3} suggests that posterior exploration is
still challenging in this problem.

The results for the simulated data (Figure~\ref{fig:masses1})
were less complex because of the definite existence of one substructure.
Its mass and position provide some information about the hyperparameters
$\boldsymbol{\alpha}_{\rm lens}$, restricting the probability that
high mass substructures exist far from the image.
